\textbf{\sc Towards Deliberative Control in Marine Robotics}

### 2012-02-24 future.tex
updates to arch and results mostly cosmetic
final updates

% Where do we see all this going esp. in the context of embedded AI Planning
% for marine robots? This can be (and should be)
% speculative. Shore-side/onboard autonomy coordination and control of
% multiple vehicles is an important result of this activity.

% 1.6 Future Work?
% explain where EUROPA is going? (ANML, incorporation of classical planning algorithms, soft constraints, better search, etc).

% The main advantage of CP is the richness of the constraints that can be modeled and the explicit use of logical inference to prune the search space

@@

Feature recognition for Sampling is a general research thrust; however
it has specific relevance in the context of \rx and AUV control. As
noted in Section \ref{sec:results} we have already demonstrated
Machine Learning techniques for recognition of select features in the
coastal ocean. A principal challenge that remains is that of obtaining
expert labeled data for any supervised or semi-supervised learning
methods.  While supervised techniques for learning classifiers such as
decision trees (\cite{Quinlan93-dtrees}), instance-based
classification (\cite{Aha-ibl-ml91}), Bayesian classifiers
(\cite{Jensen2001-BNetworks}) and artificial neural networks
(\cite{ANNsurvey-2000}) are popular, given the vast amount of data
available and poor understanding of correlation between physical and
biogeochemical variables in the coastal ocean, our work on INLs
\cite{mcgann08d,ryan10} and reliance on a oceanographic expert has
shown that scaling to different problems remains problematic. This is
a prime motivation to move towards more semi-supervised methods such
as \cite{kumar11,sergio12}. With the advent of the \od however,
another direction our research is taking us is in Recommender systems
\cite{Adomavicius05} as noted in Section \ref{sec:results}. We are
working to design and deploy software infrastructure that will learn
from user tagged data. To do so, we will use tagging techniques
similar to what users of commercial image sharing sites like Flickr
and Picassa do, to learn the context of the data (including remote
sensing). In addition, the \od will integrate statistical Machine
Learning techniques with incoming asynchronous data stream obtained by
sensors and platforms. Identification models will be based on existing
training sets for INLs and plumes; models for other features of
scientific interests driven by \can requirements will be built in
addition. Existing AUV archives for instance provide sufficient data
for a learning base for a range of features of interest to \can; in
addition we will build software that will learn to incorporate
incoming data streams to provide an existence proof of a feature of
interest.

>>>>

@@

Feature recognition for Sampling is a general research thrust; however
it has specific relevance in the context of \rx and AUV control. As
noted in Section \ref{sec:results} we have already demonstrated
Machine Learning techniques for recognition of select features in the
coastal ocean. A principal challenge that remains is that of obtaining
expert labeled data for any supervised or semi-supervised learning
methods.  While supervised techniques for learning classifiers such as
decision trees (\cite{Quinlan93-dtrees}), instance-based
classification (\cite{Aha-ibl-ml91}), Bayesian classifiers
(\cite{Jensen2001-BNetworks}) and artificial neural networks
(\cite{ANNsurvey-2000}) are popular, given the vast amount of data
available and poor understanding of correlation between physical and
biogeochemical variables in the coastal ocean, our work on INLs
\cite{mcgann08d,ryan10} and reliance on a oceanographic expert has
shown that scaling to different problems remains problematic. This is
a prime motivation to move towards more semi-supervised methods such
as \cite{kumar11,sergio12}. With the advent of the \od however,
another direction our research is taking us is in \texttt{Recommender
  Systems} \cite{Adomavicius05} as noted in Section
\ref{sec:results}. We are working to design and deploy software
infrastructure that will learn from user tagged data. To do so, we
will use tagging techniques similar to what users of commercial image
sharing sites like Flickr and Picassa do, to learn the context of the
data (including remote sensing). In addition, the \od will integrate
statistical Machine Learning techniques with incoming asynchronous
data stream obtained by sensors and platforms. Identification models
will be based on existing training sets for INLs and plumes; models
for other features of scientific interests driven by \can requirements
will be built in addition. Existing AUV archives for instance provide
sufficient data for a learning base for a range of features of
interest to \can; in addition we will build software that will learn
to incorporate incoming data streams to provide an existence proof of
a feature of interest.

A number of issues remain to be explored improving \rx and \eu
integration particularly in the context of engineering models for
execution.

Dispatchability of the {\em external} temporally flexible partial
plans continues to be an open research problem. Substantial progress
has been made \cite{vidal97,mus98a,tsam98,morris00} by analyzing the
plan structure of static offline temporal plans. However in the
context of embedded plan execution, balancing model design while
dealing with execution uncertainty requires further research. For
instance it should be possible to distinguish part of the plan that
contributes to a reactor goal, which often needs to be dispatched as
early as possible with other tokens describing world evolution with a
preference to dispatch as late as possible. Making this distinction
would improve the quality of agent behavior while avoiding extra model
complexity.

More abstractly at the architectural level, the ability to dynamically
manipulate the reactor graph would be of importance while creating new
reactors or modifying timelines. Doing so would provide more robust
failure recovery mechanisms especially after a reactor has failed to
synchronize. For example, an alternate reactor could take ownership of
timelines left invalid by the destruction of another failed reactor
ensuring a transparent implementation of component redundancy. While
the current design broadly provides for such a capability, the \rx API
has yet to implement it while ensuring other design elements within
\rx and \eu are not compromised.

Finally and importantly, integration of reasoning about resources
within \rx and tying it to existing \eu capabilities needs to be
demonstrated. Specifically \rx needs to be able to update resource
levels as incoming observations.  Doing so would ensure plan
adaptation occurs dynamically when planned and observed resource
values diverge -- for instance if the vehicle looses a bank of
batteries during the course of a mission.  Enhanced modeling
capabilities and augmented domain-independent search within \eu would
make it easier for \rx users to create such domain models and to
perform more proactive deliberation.

Along with all these potential extensions, we are also very cognizant
that the learning curve associated with \rx and \eu for a general user
in the marine robotics community needs to be tackled. At present, the
complex system that underlies \rx and its underlying framework, even
if well documented, can be intimidating for a naive first time user.

### 2012-02-24 intro.tex
final updates

Autonomous platforms in the marine environment have had a substantial
and immediate impact for both civil and military
applications. Autonomous Underwater Vehicles (AUVs) in particular have
altered the way oceanographers are able to reach beyond the surface
and make observations which are not only at higher resolution but have
altered both the discipline and impacted the engineering methodologies
behind Sampling, Control and Robotics itself
\cite{Brierley08032002,ryan05,Thomas06,Yoerger01012007,Incze2009,Rigby10}. Yet
challenges remain, principally in making such robotic devices more
attune to the environment, to sample large swaths at the Meso-scale
($> 50$ km\textsuperscript{2}) and doing so systematically and
adaptively. Further, over the last two decades substantial
improvements in hardware including propulsion, battery technologies
and hardware have not been kept abreast of in software especially in
algorithmic methods in Sampling and Control and in systematic
engineering and software engineering development.

This is one important reason that periodic calls towards sustained
exploration and sampling \cite{rudnick03} are issued by scientists for
whom the scope and scale of observations have become increasingly
tractable using autonomous platforms both below and on the
surface. The complexity of oceanic processes and the chronic
under-sampling of the coastal oceans especially of dynamic features
spread over large spatial scales (from meters to kilometers in extent)
with dynamic biological activity across the temporal spectrum (in the
order of hours to weeks), has made it challenging for robotic
platforms to autonomously track and sample. A prominent example of
such a process is coastal algal blooms which are patchy and could
cover large coastal zones. Persistent observation of such dynamic
events dictates that our robotic assets track and sample such patches
which can evolve rapidly due to inherent bio-geochemical activity,
advection as well as diffusion with the water mass it resides in.

@@

In this chapter, we propose an alternative view to decision-making
using in-situ automated Planning \cite{ghallab04} and more important
demonstrate that while earlier control paradigms primarily driven by
\emph{reactive} approaches to control can be augmented to not only
provide targeted observation and sampling, but also a more nuanced and
balanced consideration of mission objectives, environmental conditions
and available resources. Planning (or deliberation or projection in
action space) is important to balance current needs of a robot with
future desires or goals. And to do such planning in the context of
time and available resources in-situ on the robot. Planning without
time or action planning in of itself is not adequate; dealing with
time ensures that the balance between now and the future is handled
systematically. Yet, plans have to be tied to robotic action and in
the real-world which is not static. Further, plans project future
state which is not only likely to change, but is based on substantial
uncertainty\footnote{It is important to remember that these principles
  are broad much as Dwight Eisenhower is reputed to have said
  ``\emph{In preparing for battle I have always found that plans are
    useless, but planning is indispensable}'' and ``\emph{Failing to
    plan is planning to fail}''}. Modeling of the marine environment
to provide cues is hampered by poor understanding of the worlds ocean
and poor predictive skill of ocean models and often little to no
availability of a priori synoptic views of the survey area.

AUV control architectures have primarily been driven by reactive
Subsumption-based approaches \cite{brooks86}. They have been adequate
to doing routine straight-line survey tracts which have played a
valuable role in the acceptance of AUVs as a new tool for ocean
science and DoD applications. However, scaling of problem domains in
addition to dealing with event-response and discovery in the
water-column challenges \emph{how} controllers are written in the
first place, maintained and upgraded over the life-cycle of the
platform. Dealing with off-nominal endogenous or exogenous conditions
including sensor failures stretches how AUVs are currently being
utilized; one consequence for instance, is any off-nominal condition
results in a \emph{fail-safe} surface mode to radio for help. Further
these approaches, words to the contrary are not only hard to maintain,
are not 'adaptive' enough. Despite vigorous defense by Brooks early on
\cite{Brooks91intelligencewithoutrea,Brooks91intelligencewithoutrep},
autonomous systems require a model if not of the world then of the 
\emph{physics} of the vehicle when it interacts with the world.

Persistent and early legitimate criticism of deliberation on the other
hand also well articulated by Brooks
\cite{Brooks91intelligencewithoutrea,Brooks91intelligencewithoutrep}
was targeting the performance of in-situ deliberation. Terrestrial
robotic platforms were well-known to frequently stop and ``plan''; 
assuming in the meantime that the world will remain static. In
the marine robotics world, such a situation would simply not be
permissible given the need to make continuous observations without
stopping an under-articulated robotic platform such as an AUV. However
algorithmic and implementational advances by the AI community over the
years has produced a range of planners with demonstrated embedded
capabilities
\cite{simmons94,Haigh98,alami:1998p820,chien00,mus98,teichteil07}. Our
work has been influenced by and in turn influenced a wide range of
activities in the field of AI Planning and Plan Execution with
demonstrated real-world and highly visible capabilities in the Space
domain \cite{mus98,rajan00,aichang04,bresina05} for NASA. Lessons
learned have been applied to the world of marine robotics which we
bring together in this chapter. Important lessons that were derived
from this rich legacy involved the following:

@@

\item A model-based approach separating control formulation of the
  platform which describes operational characteristics from the
  domain. The controller can then be rigorously tested and certified
  even as the domain model evolves to fit diverse applications.

@@

This chapter is organized as follows. We start with some foundational
concepts in our Planning framework in Section~\ref{sec:concepts} which
includes a brief overview of AI Planning in
Section~\ref{sec:planningfound}. In Section~\ref{sec:basics} we go
deeper into the \eu framework itself, the basis of all deliberation
within \rxe. A key portion is Section~\ref{sec:europa:pr} which
details the internal plan representation essential to planning within
a situated agent like \rxe. Modeling
(Section~\ref{sec:europa:modeling}), Inference
(Section~\ref{sec:europa:inference}) and Search
(Section~\ref{sec:europa:search}) further detail the important details
associated with \eu for the and relevant to how \rx works to the
reader. With these foundational elements out of the way, we transition
to \rx itself, in Section~\ref{sec:arch}, bring out architectural
details in Section~\ref{sec:arch:trex}, deal with the execution cycle
in Section~\ref{sec:arch:exec}, highlight how \rx deliberates in
Section~\ref{sec:arch:europa} finally delving deeper into the key
concept of synchronization in Section~\ref{sec:arch:synch}. Further
details on deliberation within \rx (Section~\ref{sec:arch:plan}) and
how planning and execution are interleaved are shown in
Section~\ref{sec:arch:intertwine}. We offer results from our
experiments at sea in Section~\ref{sec:results}, offer some insights
into the near-term future goals of our research in
Section~\ref{sec:future} and conclude with
Section~\ref{sec:conclusion}.

>>>>

Autonomous platforms in the marine environment have had a substantial
and immediate impact for both civil and military
applications. Autonomous Underwater Vehicles (AUVs) have not only
altered the ability of oceanographers to reach beyond the surface and
make high resolution observations, but have impacted the engineering
methodologies behind sampling, control and robotics itself
\cite{Brierley08032002,ryan05,Thomas06,Yoerger01012007,Incze2009,Rigby10}. Yet
challenges remain, principally in making such robotic devices more
attune to the environment, to sample large swaths at the Meso-scale
($> 50$ km\textsuperscript{2}) and doing so systematically and
adaptively. Further, over the last two decades substantial
improvements in hardware including propulsion and battery technologies
have outpaced software especially in algorithmic methods in sampling
and control and in systematic software engineering development.

Large scale high resolution observations have become increasingly
tractable using autonomous platforms both on and below the surface.
This, in turn, has lead to periodic calls for sustained exploration
and sampling \cite{rudnick03}, illustrating the high scientific value
placed on such data.  The complexity of oceanic processes and the
chronic under-sampling of the coastal oceans, especially of dynamic
features studied over a variety of spatial and temporal scales, has
made it challenging for robotic platforms to autonomously track and
sample. A prominent example of such a process is coastal algal blooms
which are patchy and could cover large coastal zones. Persistent
observation of such dynamic events dictates that our robotic assets
track and sample such patches which can evolve rapidly due to inherent
bio-geochemical activity, as well as advection and diffusion of the
water mass it resides in.

@@

In this chapter, we propose an alternative view to decision-making
using in-situ automated Planning \cite{ghallab04}. More importantly,
we demonstrate that earlier control paradigms (primarily driven by
\emph{reactive} approaches to control) can be augmented to provide
targeted observation and sampling as well as a more nuanced and
balanced consideration of mission objectives, environmental conditions
and available resources. Planning (or deliberation or projection in
action space) is important to balance current needs of a robot with
future desires or goals taking into account time and available
resources in-situ on the robot. Planning without time or action
planning in and of itself is not adequate; dealing with time ensures
that the balance between now and the future is handled
systematically. Yet, plans have to be tied to robotic action and in
the real-world which is not static. Further, plans project future
state which is not only likely to change, but is based on substantial
uncertainty\footnote{It is important to remember that these principles
  are broad much as Dwight Eisenhower is reputed to have said
  ``\emph{In preparing for battle I have always found that plans are
    useless, but planning is indispensable}'' and ``\emph{Failing to
    plan is planning to fail}''}. Use of environmental cues is often
hampered by poor predicitve skill of ocean models and little to no
availability of a priori synoptic views of the survey area

AUV control architectures have primarily been driven by reactive
Subsumption-based approaches \cite{brooks86}. They have been adequate
to doing routine straight-line survey trajectories which have played a
valuable role in the acceptance of AUVs as a new tool for ocean
science and DoD applications. However, scaling of problem domains in
addition to dealing with event-response and discovery in the
water-column challenges \emph{how} controllers are written in the
first place, maintained and upgraded over the life-cycle of the
platform. Dealing with off-nominal endogenous or exogenous conditions
including sensor failures stretches how AUVs are currently being
utilized; one consequence for instance, is any off-nominal condition
results in a \emph{fail-safe} surface mode to radio for help. Software
design using these reactive approaches, not only makes it hard for
sustaining engineering development, these approaches are not
adequately adaptive for emerging ocean science problems.  Despite
vigorous defense by Brooks early on
\cite{Brooks91intelligencewithoutrea,Brooks91intelligencewithoutrep},
autonomous systems require a model if not of the world then of the
\emph{physics} of the vehicle when it interacts with the world.

Criticism of AI planning, on the other hand, also well articulated by
Brooks, was targeting the performance of in-situ
deliberation. Terrestrial robotic platforms were well-known to
frequently stop and ``plan''; assuming in the meantime that the world
will remain static. In the marine robotics world, such a situation
would simply not be permissible given the need to make continuous
observations without stopping an under-articulated robotic platform
such as an AUV. However, algorithmic and implementational advances by
the AI community over the years have produced a range of planners with
demonstrated embedded capabilities
\cite{simmons94,Haigh98,alami:1998p820,chien00,mus98,teichteil07}. Our
work has been influenced by and in turn influenced a wide range of
activities in the field of AI Planning and Plan Execution with
demonstrated real-world and highly visible capabilities in the Space
domain \cite{mus98,rajan00,aichang04,bresina05} for NASA. Lessons
learned have been applied to the world of marine robotics which we
bring together in this chapter. Important lessons that were derived
from this rich legacy included the following:

@@

\item A model-based approach should separate control formulation of
  the platform which describes operational characteristics, from the
  domain. The controller can then be rigorously tested and certified
  even as the domain model evolves to fit diverse applications.

@@

This chapter is organized as follows. We start with some foundational
concepts in our Planning framework in Section~\ref{sec:concepts} which
includes a brief overview of AI Planning in
Section~\ref{sec:planningfound}. In Section~\ref{sec:basics} we go
deeper into the \eu framework itself, the basis of all deliberation
within \rxe. A key portion is Section~\ref{sec:europa:pr} which
details the internal plan representation within a situated agent like
\rxe. Modeling (Section~\ref{sec:europa:modeling}), Inference
(Section~\ref{sec:europa:inference}) and Search
(Section~\ref{sec:europa:search}) articulates important details
associated with \eu critical to \rxe.  With these foundational
elements of deliberation out of the way, we transition to \rx itself
in Section~\ref{sec:arch}, bring out architectural details in
Section~\ref{sec:arch:trex}, deal with the execution cycle in
Section~\ref{sec:arch:exec}, highlight how \rx deliberates in
Section~\ref{sec:arch:europa}, finally delving deeper into the key
concept of synchronization in Section~\ref{sec:arch:synch}. Further
details on deliberation within \rx (Section~\ref{sec:arch:plan}) and
how planning and execution are interleaved are shown in
Section~\ref{sec:arch:intertwine}. We offer results from our
experiments at sea in Section~\ref{sec:results}, offer some insights
into the near-term future goals of our research in
Section~\ref{sec:future} and conclude with
Section~\ref{sec:conclusion}.
